---
title: "scCloneVar Workflow"
author: "Chris Chen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scCloneVar Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
  #eval = FALSE
)
```

<div style="font-size: 16px;">
## Overview

This vignette demonstrates a complete scCloneVar workflow using the included demo dataset (scCloneVar_test_demo).

The analysis proceeds through the following stages:

1. Clone size distribution profiling

2. Output Activity (OA) analysis

3. Clone-wise differential expression

4. Transcriptional heterogeneity analysis (PCA-based distances)

5. Differential variance (DVG) analysis

6. GSEA on ranked heterogeneity statistics
</div>

## Load Package and Demo Data 

```{r}
library(scCloneVar)
library(Seurat)

data(scCloneVar_test_demo)

seurat_obj <- scCloneVar_test_demo
```


```{r}
seurat_obj
```

<div style="font-size: 16px;">

## 1. Clone Size Distribution

This step visualizes clone expansion patterns and computes descriptive statistics across samples.

Clonal expansion is a fundamental property of hematopoietic dynamics, reflecting the proliferative fitness and competitive advantage of individual clones within a sample. Quantifying relative clone size allows comparison across replicates and conditions while controlling for differences in total cell number.


For each replicate, relative clone frequency is defined as:

Relative clone frequency is calculated as  
</div>

<div style="font-size: 20px;">

<br>

$$
\mathrm{RelFreq} =
\frac{\text{cells in clone}}{\text{total cells in replicate}}
$$

<br>

</div>

<div style="font-size: 16px;">
Clones below a user-defined frequency threshold are grouped as Other, enabling clearer visualization of dominant expansions.
</div>

```{r,fig.width=4, fig.height=2.5, out.width="70%", dpi=300}
comparison_list <- list(
  list(seurat_obj = seurat_obj,samples = grep("^Y_vitro", unique(seurat_obj$sampleName), value = TRUE),label_counts = data.frame(
      Sample = grep("^Y_vitro", unique(seurat_obj$sampleName), value = TRUE),
      Replicate = unique(seurat_obj$Rep[grepl("^Y_vitro", seurat_obj$sampleName)]),
      SampleLabel = grep("^Y_vitro", unique(seurat_obj$sampleName), value = TRUE)),
    threshold = 0.05,
    title = "Relative Clone Size Distribution: Y_vitro",
    name = "Y_vitro"),
  list(
    seurat_obj = seurat_obj,
    samples = grep("^O_vitro", unique(seurat_obj$sampleName), value = TRUE),
    label_counts = data.frame(
      Sample = grep("^O_vitro", unique(seurat_obj$sampleName), value = TRUE),
      Replicate = unique(seurat_obj$Rep[grepl("^O_vitro", seurat_obj$sampleName)]),
      SampleLabel = grep("^O_vitro", unique(seurat_obj$sampleName), value = TRUE)),
    threshold = 0.05,
    title = "Relative Clone Size Distribution: O_vitro",
    name = "O_vitro"))

dist_out <- run_clone_distribution_engine(comparison_list)

dist_out$plots$Y_vitro
dist_out$plots$O_vitro
```
<div style="font-size: 16px;">
Relative clonal size summary table is also provided:
</div>
```{r}
dist_out$descriptive_stats$Y_vitro
```
<div style="font-size: 16px;">
## 2. Output Activity (OA) Analysis

This step quantifies clone-level output bias and classifies clones into low and high output groups.

Output Activity (OA) measures the differentiation bias of each clone by comparing its contribution to non-HSC versus HSC compartments. Following the conceptual framework described by Singh et al., Cell Stem Cell (2025) https://www.cell.com/cell-stem-cell/fulltext/S1934-5909(25)00012-8, OA reflects the functional output potential of a clone — that is, whether a clone preferentially produces differentiated progeny or remains stem-biased. In this implementation, OA is computed as the relative enrichment of non-HSC output normalized to HSC contribution within each clone.


For a given clone $c$, Output Activity is defined as:

$$
\mathrm{OA}_c = 
\frac{\mathrm{nonHSC\_freq}_c + \epsilon}
     {\mathrm{HSC\_freq}_c + \epsilon}
$$

where

$$
\mathrm{HSC\_freq}_c = 
\frac{N^{\mathrm{HSC}}_c}
     {N^{\mathrm{HSC}}_c + N^{\mathrm{nonHSC}}_c}
$$

$$
\mathrm{nonHSC\_freq}_c = 
\frac{N^{\mathrm{nonHSC}}_c}
     {N^{\mathrm{HSC}}_c + N^{\mathrm{nonHSC}}_c}
$$

and

$$
\epsilon = 10^{-6}
$$


Higher OA values indicate clones with greater differentiation output potential, whereas lower OA values represent stem-biased clones.

### Required Inputs

The function requires a Seurat object containing:
-A UMAP reduction named "umap",
-Metadata columns specifying celltype and clone identity (default "CloneID"),
-Defined HSC cell types (default: LT-HSC, ST-HSC).
-Optional arguments allow specification of non-HSC cell types, DEG cell subsets, reference marker lists, color mappings, and statistical thresholds.
</div>

```{r}
# Define low and high output marker 

low_output_markers <- c(
  "Mpl","Ifitm3","Ifitm1","Tgm2","H2-K1","Socs2","Mycn","Nupr1","Hacd4",
  "Mllt3","Gda","B2m","Procr","Txnip","Clu","Sult1a1","S100a6","Rpl36a",
  "Tsc22d1","Gng11","Ccnd2","mt-Co3","Tmem176b","Lmo2","Ly6a","Mmrn1",
  "Trim47","St3gal1","Mecom","Pik3r1","Adgrg1","Esam","Ryk","Rps21","Hoxb8",
  "Ccnd1","Uba7","Rps28","Serpina3g","Rpl37a","Fkbp1a","Pdzk1ip1","Selp",
  "Eif4a2","Tmem176a","Bex4","Grb10","Jam3","Ptk2b","Gimap8","Csgalnact1",
  "H2-Q7","App","Mylk","Casp12","Rhof","Laptm4a","Tcf15","Gabarapl1","Ctsl",
  "Bex1","Rpl21","D630039A03Rik","Myof","Glul","Ppic","Tpt1","Rpl5","Cdkn1c",
  "mt-Nd5","Abcg3","Aplp2","Art4","Tbxas1","Cish","Vwf","Scarf1","mt-Atp6",
  "H2-Eb1","H2-D1","Cd74","Iigp1","Aldh1a1","Gstm1","mt-Nd4","mt-Cytb","mt-Co1"
)
high_output_markers <- c(
  "Plac8","H2afy","Cdk6","Cd34","Nkg7","Ptma","Mpo","Cd48","Stmn1","Fam117a",
  "Slc22a3","Adgrg3","Ppia","Car1","Ctsg","Flt3","Lgals1","Muc4","Gpx1",
  "Hmgb2","Ndufa4","Serpinb1a","Ccl9","Oaz1","H2afz","Crip1","Mcm7","Cpa3",
  "Vim","Ybx1","Sell","Sh3bgrl3","H3f3a","Dut","Atpif1","Ran","Hnrnpa2b1",
  "Hdgf","Mcm4","Elane","Rabgap1l","Cmtm7","Rpsa","Mcm6","Plek","Set","Atp5g3",
  "Myc","Taldo1","Tuba1b","Cks2","Slc25a5","Fam65a","Cebpa","Tmsb10","Cd52",
  "Klf1","Anp32b","Hn1","Parvg","Ffar2","Bex6","Emilin2","Itgal","Cox5a",
  "Hnrnpab","Ighm","BC035044","Lmnb1","Golm1","Bin1","Igfbp4","Tyrobp","E2f8",
  "Banf1","Cst7","Sh2d5","Aqp1","Ptbp3","Snrpf","Rfc2","Ramp1","Hmgn5","Nme1"
)


```


```{r, warning=FALSE}
## Define celltype colors manually ##
celltype_colors <- c(
  ## HSCs 
  "LT-HSC" = "#E76F51",
  "ST-HSC" = "#F4A261",
  ## Early progenitors 
  "MPP"  = "#5BC374",
  "MPP3" = "#4CAF50",
  "MPP4" = "#66CDAA",
  "CMP"  = "#C49A6C",
  "EryP" = "#F28E2B",
  "LMPP" = "#4DB6AC",
  ## Myeloid differentiation
  "GMP"   = "#1FA187",
  "GMP-3" = "#2BBBAD",
  "MDP"   = "#7C9A3D",
  "Granulocyte" = "#6C8EBF",
  ## Megakaryocyte / Platelet
  "MkP"   = "#5FBAC8",
  "MKP"   = "#4DA8B5",
  "MKP-1" = "#78C2D0",
  ## Erythroid / Megakaryocyte
  "MEP"     = "#8C973E",
  "MEP-1"   = "#7A8735",
  "MEP-2"   = "#9AAE4C",
  "MEP/EryP"= "#A4B95A",
  ## Mast / Basophil 
  "Ba/mast"   = "#C77DFF",
  "Mast-cell" = "#B565D9",
  ## Immune / Other 
  "DC"  = "#9E6E4A",
  "Mac" = "#7F5539",
  "Mast-Cells" = "#D98BB8",
  ## T cell 
  "T-cells" = "#6CA3D1",
  "T-cell"  = "#4F8AC9",
  "T-Cell"  = "#83B8E6",
  ## Undefined
  "UN" = "grey65")

## Run OA function ##

oa_out <- run_low_high_OA_analysis_for_a_single_sample(seurat_obj = seurat_obj,
  sample_label = "DemoSample",celltype_colors = celltype_colors,low_output_markers = low_output_markers,
  high_output_markers = high_output_markers, full_ref_deg_genes_list = NULL)

```

<div style="font-size: 16px;">
### Expected Outputs

The function returns a structured results object (also saved to disk) containing:
-Per-clone OA values and low/high classifications,
-Cell-level OA annotations,
-Clone-wise DEG results (via run_clonewise_DEG_suite()),
-UMAP visualizations (cell type, OA gradient, contour),
-Volcano plots and universe-restricted enrichment tests,
-A complete RDS object consolidating tables and plots.
</div>


## 3. Clone-wise Differential Expression

Differential expression analysis between low and high OA clones.

```{r,eval=FALSE}

deg_out <- run_clonewise_DEG_suite(
  seurat_obj = seurat_obj,
  clone_set1_ids = oa_out$low_clones,
  clone_set2_ids = oa_out$high_clones
)

deg_out$volcano_plot

```

## 4. PCA-Based Clone Heterogeneity

Quantifies intra- and inter-clonal transcriptional distances in PCA space.

```{r,eval=FALSE}

het_out <- compute_pca_clone_heterogeneity(
  seurat_obj = seurat_obj
)

head(het_out$clone_level_stats)
```

## 5. Differential Variance and CV Analysis

Identifies Differential Variance Genes (DVGs) across groups.

```{r,eval=FALSE}

dvg_out <- compute_variance_and_cv(
  seurat_obj = seurat_obj,
  group_var = "donor_age"
)

head(dvg_out$results)

```

## 6. Visualization of Heterogeneity Patterns

```{r,eval=FALSE}

plot_variance_cv_summaries(dvg_out$results)

```


## 7. GSEA Across MSigDB Collections

Performs enrichment analysis on ranked heterogeneity statistics.

```{r,eval=FALSE}

gsea_out <- run_msigdb_gsea_all_collections(
  hsc_results = dvg_out$results,
  rank_var = "mean_adjusted_var"
)

head(gsea_out$H)

```



